---
title: "QC On Trimmed Data"
date: "`r format(Sys.Date(), '%d %B, %Y')`"
---

```{r setup}
knitr::opts_chunk$set(
	message = FALSE, warning = FALSE,
	fig.height = 6, fig.width = 10
)
```

```{r packages}
library(ngsReports)
library(tidyverse)
library(yaml)
library(scales)
library(pander)
library(glue)
library(cowplot)
library(plotly)
```
```{r options}
panderOptions("table.split.table", Inf)
panderOptions("big.mark", ",")
theme_set(theme_bw())
suffix <- "_L001_R1.fastq.gz"
pattern <- paste0("_CB2YGANXX_.+fastq.gz")
sp <- "Rnorvegicus"
```


```{r samples}
samples <- "data/targets.csv" %>%
  here::here() %>%
  read_csv() %>% 
	mutate(
	  Filename = paste0(File, suffix)
	)
```


```{r group_cols}
group_cols <- hcl.colors(
  n = length(unique(samples$group)), 
  palette = "Zissou 1"
  ) %>%
  setNames(unique(samples$group))
```



# Quality Assessment on Trimmed Data

In the workflow, trimming was performed using the tool `AdapterRemoval` with the settings:

- Minimum length after trimming: 50
- Minimum quality score to retain: 30

## Overall Summary

```{r bothFqc}
rawFqc <- here::here("data/0_rawData/FastQC") %>%
  list.files(pattern = "zip", full.names = TRUE) %>%
  FastqcDataList() %>%
  .[fqName(.) %in% samples$Filename]
trimFqc <- here::here("data/1_trimmedData/FastQC") %>%
  list.files(pattern = "zip", full.names = TRUE) %>%
  FastqcDataList() %>%
  .[fqName(.) %in% samples$Filename]
```

After trimming, the library showing the highest level of possible adapter content contained `r getModule(trimFqc, "Adapter") %>% pivot_longer(-any_of(c("Filename", "Position"))) %>% group_by(Filename, Position) %>% summarise(Total = sum(value)) %>% summarise(max = max(Total)) %>% dplyr::filter(max == max(max)) %>% pull(max) %>% percent(accuracy = 0.01)` of reads as containing possible adapter sequences.


```{r plotSummary, fig.cap = "*Comparison of FastQC summaries before and after trimming*"}
bind_rows(
  getSummary(rawFqc) %>% mutate(data = "Raw"),
  getSummary(trimFqc) %>% mutate(data = "Trimmed"),
) %>% 
  left_join(samples) %>% 
  ggplot(aes(Category, Rat, fill = Status)) +
  geom_tile() +
  facet_grid(data~.) +
  scale_fill_manual(values = getColours(pwf)[c("PASS", "WARN", "FAIL")]) +
  scale_x_discrete(expand = expansion(c(0, 0))) +
  scale_y_discrete(expand = expansion(c(0, 0))) +
  theme(
    axis.text.x = element_text(angle= 90, hjust = 1, vjust = 0.5),
    strip.text.y = element_text(angle = 0)
  )
```

## Library Sizes

```{r compareLibSizes}
readTotals(rawFqc) %>%
  rename(Raw = Total_Sequences) %>%
  left_join(
    readTotals(trimFqc) %>%
      rename(Trimmed = Total_Sequences)
  ) %>%
  mutate(
    Remaining = Trimmed / Raw,
    Filename = str_remove_all(Filename, suffix)
  ) %>%
  summarise(
    across(c(Remaining, Trimmed), list(min = min, mean = mean, max = max))
  ) %>%
  pivot_longer(everything()) %>%
  separate(
    name, into = c("Type", "Summary Statistic")
  ) %>%
  pivot_wider(names_from = Type, values_from = value) %>%
  mutate(
    Remaining = percent(Remaining, accuracy = 0.1),
    `Summary Statistic` = str_to_title(`Summary Statistic`)
  ) %>%
  rename(Reads = Trimmed) %>%
  pander(
    caption = "*Summary statistics showing the results after trimming*"
  )
```

```{r plot-lib-sizes, fig.cap = "Comparison of library sizes before and after trimming."}
readTotals(rawFqc) %>%
  rename(Raw = Total_Sequences) %>%
  left_join(
    readTotals(trimFqc) %>%
      rename(Retained = Total_Sequences),
    by = "Filename"
  ) %>% 
  left_join(samples, by = "Filename")%>% 
  mutate(Discarded = Raw - Retained) %>% 
  pivot_longer(
    cols = all_of(c("Retained", "Discarded")),
    names_to = "Status",
    values_to = "Reads"
  ) %>% 
  ggplot(aes(fct_rev(Rat), Reads/1e6, fill = Status)) + 
  geom_col() +
  facet_grid(group~. , scales = "free_y", space = "free_y") +
  scale_y_continuous(expand = expansion(c(0, 0.05))) +
  scale_fill_viridis_d(option = "cividis", direction = -1) +
  labs(
    x = "Sample", y = "Reads (millions)"
  ) +
  coord_flip()
```


## Sequence Length Distribution

```{r plotLengthDist, fig.cap="*Distribution of read lengths after trimming*"}
ggplotly(
  getModule(trimFqc, "Sequence_Length") %>%
    group_by(Filename) %>%
    mutate(
      `Cumulative Total` = cumsum(Count),
      `Cumulative Percent` = percent(`Cumulative Total` / max(`Cumulative Total`))
    ) %>%
    ungroup() %>%
    left_join(samples, by = "Filename") %>%
    rename_all(str_to_title) %>% 
  arrange(Lower) %>% 
  mutate(Length = fct_inorder(Length)) %>%
    ggplot(aes(Length, `Cumulative Total`, group = Filename, label = `Cumulative Percent`)) +
    geom_line(aes(colour = Group), size = 1/4) +
    scale_y_continuous(label = comma) +
    scale_colour_manual(
      values = group_cols
    ) 
)
```

## GC Content


```{r plotGC, fig.cap=glue("*GC content shown as the % above and below the theoretical GC content for the {sp} transcriptome. The peaks at 62% and 71% are strongly suggestive of incomplete rRNA removal*")}
plotGcContent(
  x = trimFqc, 
  pattern = pattern, 
  species = sp, 
  plotType = "line",
  gcType = "Trans",
  usePlotly = TRUE,
  dendrogram = TRUE,
  cluster = TRUE
  )
```


## Sequence Content

```{r plotSeqContentResid, fig.cap = "*Base and Position specific residuals for each sample. The mean base content at each position was calculated for each nucleotide, and the sample-specific residuals calculated.*"}
plotly::ggplotly(
  getModule(trimFqc, module = "Per_base_sequence_content") %>% 
    mutate(Base = fct_inorder(Base)) %>%
    group_by(Base) %>% 
    mutate(
      across(c("A", "C", "G", "T"), function(x){x - mean(x)}) 
    ) %>% 
    pivot_longer(
      cols = c("A", "C", "G", "T"), 
      names_to = "Nuc", 
      values_to = "resid"
    ) %>%
    left_join(samples, by = "Filename") %>%
    ggplot(
      aes(Base, resid, group = Filename, colour = group)
    ) + 
    geom_line() +
    facet_wrap(~Nuc) + 
    scale_colour_manual(values = group_cols) +
    labs(
      x = "Read Position", y = "Residual", colour = "Group"
    ) +
    theme(
      axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)
    )
)
```

<button type="button" class="btn btn-default btn-sessioninfo" data-toggle="collapse" data-target="#sessioninfo" style="display: block;">
<span class="glyphicon glyphicon-info-sign" aria-hidden="true"></span> Session information
</button>
</p>
<div id="sessioninfo" class="collapse">
```{r session-info, echo=FALSE}
pander::pander(sessionInfo())
```
</div>